---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).


### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

```{r}
#Print will take a string and return it in the terminal 
print("Hello World")
```
```{r}
#The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains.
library(tidyverse)
```

```{r}
#The number function saves the number as a variable in the environment. 
number <- 2 
```

```{r}
#glimpse() function gives glimpse of data 
```

```{r}
#head() function analyzes certain columns of data  
```