---
title: "Data Analysis Project"
author: "Sandra Smith, Lauren Bentley, Katelynn Winebrenner"
date: "May 13, 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction, Preliminary Analysis and Definitions

In this notebook, we are analyzing Maryland precinct data from the 2020 and 2024 elections. To do this, we have loaded in 2 data sets from the Maryland State Board of Elections. 

The 2020 data set has 23 columns and 95283 rows. 
The 2024 data set has 21 columns and 117875 rows. 
Both data sets show each precinct, information about the precinct, each race, each candidate, the winner, how many votes each candidate received and method of vote. 

Because each year has different candidates and different races, we should identify which races happened both years and in all counties in order to make a clear analysis. We should also identify clear candidates in each race so that we aren't analyzing someone who is listed, but received one write in vote, for example.

To meet these parameters, we will focus on the race for president because it happened both years and in every precinct. In terms of candidates, we will focus on Donald Trump and Joe Biden for 2020 and Kamala Harris and Donald Trump for 2024.

We should be able to answer most of our questions with this data. However, we will need to load in data from the census in order to answer questions about economic status and identify. I don't know how easy that will be to do by precinct, but we'll see. We'll also need to include information about eligible voters and registered voters to determine turnout.


## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
options(scipen=999)
library(tidyverse)
library(lubridate)
library(tidycensus)
library(janitor)
library(sf)
```

## Load and Clean Data
```{r}
# Load required data
uncleaned_precincts2020 <- read_csv("data/All_By_Precinct_2020_General.csv")
uncleaned_precincts2024 <- read_csv("data/PG24_AllPrecincts.csv")
```

```{r}
# Clean 2020, column names to match, county names, combine column district and column precinct, select columns
precincts2020 <- uncleaned_precincts2020 |>
  clean_names() |>
  mutate(county_name = case_when(
    county == "01" ~ "Allegany County",
    county == "02" ~ "Anne Arundel County",
    county == "03" ~ "Baltimore City",
    county == "04" ~ "Baltimore County",
    county == "05" ~ "Calvert County",
    county == "06" ~ "Caroline County",
    county == "07" ~ "Carroll County",
    county == "08" ~ "Cecil City",
    county == "09" ~ "Charles County",
    county == "10" ~ "Dorchester County",
    county == "11" ~ "Frederick County",
    county == "12" ~ "Garrett County",
    county == "13" ~ "Harford County",
    county == "14" ~ "Howard County",
    county == "15" ~ "Kent County",
    county == "16" ~ "Montgomery County",
    county == "17" ~ "Prince George's County",
    county == "18" ~ "Queen Anne's City",
    county == "19" ~ "Saint Mary's County",
    county == "20" ~ "Somerset County",
    county == "21" ~ "Talbot County",
    county == "22" ~ "Washington County",
    county == "23" ~ "Wicomico County",
    county == "24" ~ "Worcester County"
  ))|>
  mutate(election_district_precinct = str_c(election_district, election_precinct, sep = "-"),
         year = 2020)|> #chatgpt, str_c function to combine two column outputs by - in single column
  rename(
    election_day = election_day_votes,
    early= early_voting_votes,
    mail_in_1= by_mail_votes,
    mail_in_2 = by_mail_2_votes,
    provisional= prov_votes,
    total= total_votes,
    candidate_name = candidate_name,
    party = party
         )|>
      select(year,county, county_name, election_district_precinct, office_name, candidate_name, party, election_day, early, mail_in_1, mail_in_2, provisional, total)|>
  filter(
    office_name == "President - Vice Pres", 
    candidate_name %in% c("Joe Biden", "Donald J. Trump")) #%n% from chatgpt


# Clean 2024, column names to match, create total votes, select columns
precincts2024 <- uncleaned_precincts2024|>
  clean_names()|>
   rename(
     election_day = election_night_votes,
     mail_in_1 = mail_in_ballot_1_votes,
     mail_in_2 = mail_in_ballot_2_votes,
     early = early_votes,
     provisional= provisional_votes,
     candidate_name = candidate_name,
     party = party
     )|>
  mutate(
    total = (election_day + mail_in_1 + mail_in_2 + early + provisional),
    year = 2024
  )|>
  
    select(year, county, county_name, election_district_precinct, office_name, candidate_name, party, election_day, early, mail_in_1, mail_in_2, provisional, total)|>
 filter(
    office_name == "President - Vice Pres",
    candidate_name %in% c("Kamala D. Harris and Tim Walz", "Donald J. Trump and JD Vance") #%n% from chatgpt
  )

```

```{r}
# combine dataframes, not joining, but placing them on top of one another
precincts_combined <- bind_rows(precincts2020, precincts2024)

```


## Questions to answer
In this notebook, we are answering five questions about the data.

### Question 1

* **Question**: Which precincts experienced the most significant change in party support? 

* **Analysis summary**: To find out which precincts experienced the most significant change in party support, we filtered each dataset with 2020 and 2024 precincts to show only the precinct, county and party. Then, we joined the datasets together and calculated the democratic and republican percent changes for each. Through this analysis, we found that Baltimore County 009-029 had the largest republican percent change from 2020 to 2024 and Baltimore County 009-027 also had the largest democratic percent change from 2020 to 2024. Washington County and Harford County precincts also had large percent change in party support. Prince George's County 019-005 had the lowest democratic percent change and Washington County 013-002 had the lowest republican percent change. 

```{r}
#Summarize votes by party (2020)
votes_2020 <- precincts_combined |>
  filter(year == 2020) |>
  group_by(election_district_precinct, county_name, party) |>
  summarize(total_votes_2020 = sum(total,na.rm = TRUE), .groups= "drop")

#Summarize votes by party (2024)
votes_2024 <- precincts_combined |>
  filter(year == 2024) |>
  group_by(election_district_precinct, county_name, party) |>
  summarize(total_votes_2024 = sum(total, na.rm=TRUE), .groups="drop")

#Join 2020 and 2024 datasets by party and precinct 
vote_changes <- left_join(votes_2020, votes_2024, by = c("election_district_precinct", "county_name", "party"))

#Democrat percent change 
democratic_changes <- vote_changes |>
  filter(party == "DEM") |>
  mutate(
    percent_change = ((total_votes_2024 - total_votes_2020) / total_votes_2020) * 100
  ) |>
  select(election_district_precinct, county_name, percent_change) |>
  arrange(desc(percent_change))

#Republican percent change 
republican_changes <- vote_changes |>
  filter(party == "REP") |>
  mutate(
    percent_change = ((total_votes_2024 - total_votes_2020) / total_votes_2020) * 100 
  ) |>
  select(election_district_precinct, county_name, percent_change) |>
  arrange(desc(percent_change))
```






### Question 2

* **Question**: How did voter turnout vary across precincts? What factors contributed to this? 

* **Analysis summary**: To find how voter turnout varied across precincts from 2020 to 2024, we calculated the total number of votes by precinct for each year. We then seperated these datasets by year and joined them by precinct. We then calculated the percent changes for both increasing and decreasing votes. From this, we only looked at the top 10 precincts from each election district. We found that Prince George's County and Washington County had the largest decrease in voter turnout from 2020 to 2024. Prince George's County had small numbers within the dataset, with only 3 voters shown -- so this number is not reflective of the other precinct changes. Harford County had the largest increase in voter turnout, along with various precincts in Baltimore County. 

```{r}
#Total votes by precinct (2020 and 2024)
voter_turnout_summary <- precincts_combined |>
  group_by(year, election_district_precinct, county_name) |>
  summarize(total_voter_turnout = sum(total,na.rm=TRUE), .groups="drop")

#Seperate by year 
voter_turnout_2020 <- voter_turnout_summary |>
  filter(year == 2020) |>
  select(election_district_precinct, county_name, total_turnout_2020 = total_voter_turnout)

voter_turnout_2024 <- voter_turnout_summary |>
  filter(year == 2024) |>
  select(election_district_precinct, county_name, total_turnout_2024 = total_voter_turnout)

#Join by precinct 
voter_turnout_changes <- left_join(voter_turnout_2020,voter_turnout_2024, by = c("election_district_precinct", "county_name")) |>
  mutate(change = total_turnout_2024 - total_turnout_2020, 
         percent_change= (change/total_turnout_2020) * 100)

#View largest increase and decrease 
top_voter_increases <- voter_turnout_changes |>
  arrange(desc(percent_change)) |>
  slice_head(n=10)

top_voter_decreases <- voter_turnout_changes |>
  arrange(percent_change) |>
  slice_head(n=10)
  


```





### Question 3

* **Question**: How did early voting and mail-in ballots influence precinct-level results? 


* **Analysis summary**: 

```{r}
#summarize votes by method in 2020
methods_2020 <- precincts_combined |>
  filter(year == 2020) |>
  group_by(
    election_district_precinct, county_name, party, early, mail_in_1, mail_in_2, provisional, total
  ) |>
    summarize(
      total_votes_20 = sum(total,na.rm = TRUE),
      .groups = "drop"
    ) |>
  mutate(year=2020)
  

#summarize votes by method in 2024
methods_2024 <- precincts_combined |>
  group_by(
    election_district_precinct, county_name, party, early, mail_in_1, mail_in_2, provisional, total
  ) |>
  summarize(
    total_votes_24 = sum(total,na.rm = TRUE),
    .groups = "drop"
    ) |>
  mutate(year=2024)

methods_combined <- bind_rows (methods_2020, methods_2024)


#calculate percentage difference
methods_combined <- methods_combined |>
  mutate(
    early_percent = early/total,
    mail_in_percent = (mail_in_1 + mail_in_2) / total,
    provisional_percent = provisional / total
  )

methods_combined <- methods_combined |>
  group_by(year, party) |>
  summarise(
    average_early = mean(early_percent, na.rm = TRUE),
    average_mail = mean(mail_in_percent, na.rm = TRUE),
    average_provisional = mean(provisional_percent, na.rm = TRUE),
    .groups = "drop"
  )
  
print(methods_combined)

```





### Question 4

* **Question**: How did voting behavior differ between precincts based on economic status? 

* **Analysis summary**: 
```{r}

precincts_shape <- st_read("data/Maryland_Election_Boundaries_-_Precincts_2022/Maryland_Election_Boundaries_-_Precincts_2022.shp")|>
  clean_names()|>
   mutate(
    name = str_extract(name, "\\d{1,2}-\\d{3}"),  # Extract the district-precinct code (e.g., "29-002")
    district = str_extract(name, "^\\d{1,2}"),  # Extract the district number (e.g., "29", "5", "12")
    precinct_code = str_pad(district, width = 3, side = "left", pad = "0"),  # Pad district with leading zeros
    name = str_replace(name, "^\\d{1,2}", precinct_code)  # Replace original district with padded district
  ) |>
rename(election_district_precinct = name)


#join precinct data to geo data

geometry_precincts_2020 <- precincts_shape |>
  left_join(precincts2020, join_by(election_district_precinct)) 

geometry_precincts_2024 <- precincts_shape |>
  left_join(precincts2024, join_by(election_district_precinct)) 

#load census tracts, income data
median_income_pop <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "MD",
  year = 2021,
  geometry = TRUE,
  ) |> 
  clean_names()
  
  

median_income_pop |>
  as.data.frame()



```

```{r error=TRUE}
geometry_precincts_2020 <- geometry_precincts_2020 |>
  st_transform(crs=5070)

geometry_precincts_2024 <- geometry_precincts_2024 |>
  st_transform(crs=5070)

median_income_pop <- median_income_pop|>
  st_transform(crs=5070)


# Perform a spatial join using st_within
#spatial_join_2020 <- st_within(geometry_precincts_2020, median_income_pop, sparse = FALSE)

#spatial_join_2024  <- st_within(geometry_precincts_2024, median_income_pop, sparse = FALSE)

```

```{r}
MAP :()
map <- mapview(geometry_precincts_2020, 
               col.region = "blue", + # Set the color of the precincts
       mapview(median_income_pop, 
                     col.region = "red", # Set the color of the census tracts
                     alpha.region = 0.8,
                     cex = 2, # Set the transparency
                     legend = TRUE) # Enable the legend

view(map)



```





### Question 5

* **Question**: How did voting behavior differ between precincts based on race, gender and age group? 

* **Analysis summary**: 
```{r}
# get census income data by tract
md_income_data <- get_acs(
  geography = "tract",
  variables = "B19013_001",
  state = "MD",
  year = 2021,
  geometry = TRUE
)
glimpse(md_income_data)

```
```{r}
#read precinct shapefile 
glimpse(md_income_data)

uncleaned_precincts_shape <- st_read("data/Maryland_Election_Boundaries_-_Precincts_2022/Maryland_Election_Boundaries_-_Precincts_2022.shp")

```
```{r}
precincts_shape <- uncleaned_precincts_shape |>
  clean_names()|>
   mutate(
    name = str_extract(name, "\\d{1,2}-\\d{3}"),  # Extract the district-precinct code (e.g., "29-002")
    district = str_extract(name, "^\\d{1,2}"),  # Extract the district number (e.g., "29", "5", "12")
    precinct_code = str_pad(district, width = 3, side = "left", pad = "0"),  # Pad district with leading zeros
    name = str_replace(name, "^\\d{1,2}", precinct_code)  # Replace original district with padded district
  ) |>
  select(name, geometry)|>  # Keep only the formatted precinct_column
rename(election_district_precinct = name)

```
```{r}
#join precinct spatial data to election data
geometry_precincts_combined <- precincts_shape |>
  left_join(precincts_combined, join_by(election_district_precinct)) 
```

```{r}
#join last dataframe with maryland income dataframe
#make same type
md_income_shape <- st_transform(md_income_data, st_crs(geometry_precincts_combined))
```
```{r}
#join precinct data to county spatial data

joined_data <- st_join(precincts_shape, md_income_shape)|>
 rename(median_income = estimate)
```

```{r}

election_economic_analysis <- left_join(joined_data, precincts_combined, by = "election_district_precinct")
```



